\begin{acknowledgments}
I would like to express my deepest appreciation to my advisors, Professors Hui Yang, Arno Puder, and Anagha Kulkarni. Their patience, support, and confidence in me were invaluable throughout my graduate journey at SFSU. I also owe a special debt of gratitude to Professor Tao He, whose crucial suggestion to incorporate a global similarity metric with the embedding vectors significantly improved the model's classification performance.

This research was made possible through the contributions of many individuals. I'd like to thank the Program Pathways Mapper (PPM) team, which includes representatives from the Kern Community College District, the Foundation for California Community Colleges, and the California Community Colleges Chancellor's Office, for their partnership in providing the foundational data for this work.  I'd like to recognize Natalie Yam, Parth Panchal, and Joanne Park for their assistance with data collection, compilation, and preliminary analysis.

This work was supported in part through the Platform for Open Learning, Academic Research, \& Innovative Scientific computing (POLARIS) High-Performance Computing (HPC) cluster at San Francisco State University. We acknowledge these resources, services, and the support provided by the Academic Technology Systems Team.

On a personal note, I am immensely thankful for my dear friends who have been there for me through thick and thin, cheering me on whenever I felt I could not continue. To my good friend, Phil, I offer my deepest appreciation for his incredible generosity and support, which made navigating the financial challenges of graduate school possible. To Julie and Bita, I am profoundly thankful for their invaluable guidance and perspective, which were instrumental to my personal growth and well-being throughout this journey.  To my brother Nick, I am eternally grateful for his quiet strength and unwavering support during times when distance and understanding meant everything.

To all who stood by me with patience, kindness, and belief---I carry your support with deep gratitude.
\end{acknowledgments}